---
title: "Free mediation models with pamlj"
author: "Marcello Gallucci"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Free mediation models with pamlj}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
library(pamlj)
```

The jamovi module `pamlj` offers three ways to set up a mediation power
analysis: the **simple** model (3 variables), the **complex** templates
(two/three parallel mediators, serial mediation) and the **free** interface,
which lets you describe _any_ recursive mediation model by typing its equations
in a model syntax. This vignette shows how to use the free interface from R, via
`pamlj::pamlmed()` with `mode = "medmodels"`.

The free interface is useful when your model does not match one of the
predefined templates: several independent variables, an arbitrary number of
mediators, serial and parallel pathways combined, and so on. From the equations,
`pamlj` builds the path model, finds every indirect (mediated) effect, and
computes the power parameters for each of them.

In R, the model is passed as a single string to the `code` argument (one
equation per line, separated by `\n`). Every other argument — `aim`, `n`,
`power`, `test`, `sig.level`, … — works as in the other mediation modes. The
results table is returned in `obj$powertab$asDF`.

# Model syntax

The model is described with one regression equation per _endogenous_ variable
(every variable that is predicted by others), written in the familiar
`lm()`/`lavaan` style. The classical simple mediation reads:

```{r}
res <- pamlj::pamlmed(
  mode  = "medmodels",
  code  = "m ~ .3*x
           y ~ .2*x + .4*m",
  aim   = "n",
  power = .80,
  test  = "joint"
)
res$powertab$asDF
```

Each line reads as _outcome `~` predictors_. Here `x` predicts the mediator `m`
with a standardized coefficient of `.3`, and `y` is predicted by both `x`
(direct effect `c' = .2`) and `m` (`b = .4`). The mediated effect `x → m → y`
is the product of the path coefficients, `.3 * .4 = .12` (the `es` / `ME`
column), and the table reports the sample size required to detect it with power
`.80`.

A few rules apply to the syntax:

* **Every predictor needs a numeric coefficient**, passed as `value*variable`
  (e.g. `.4*m`). Coefficients are completely standardized path coefficients and
  their absolute value must lie between `.001` and `.99`.
* **One equation per outcome.** A variable can appear on the left-hand side of
  at most one equation; list all its predictors on that single line.
* **No intercepts, interactions or categorical terms.** The model works on
  standardized coefficients between single variables, so intercept tokens (`1`)
  are ignored and interaction (`x:z`) or categorical (`[...]`) terms are not
  allowed.
* **The model must be a proper mediation model**: it has to be _recursive_ (no
  feedback loops) and contain at least one indirect path (a variable that is
  both an outcome and a predictor). If these conditions are not met, the
  function reports what is missing instead of running.

# Indirect effects

`pamlj` automatically enumerates every indirect effect in the model — each
simple directed path of at least two edges that runs from an exogenous variable
(a _source_, never predicted by others) to a final outcome (a _sink_, that
predicts nothing). Each such effect becomes one row of the results table,
labelled with its path in the `effect` column.

A model with two parallel mediators therefore yields two indirect effects:

```{r}
res <- pamlj::pamlmed(
  mode  = "medmodels",
  code  = "m1 ~ .4*x
           m2 ~ .5*x
           y  ~ .1*x + .2*m1 + .3*m2",
  aim   = "power",
  n     = 200,
  test  = "joint"
)
res$powertab$asDF[, c("effect", "a", "b", "es", "power")]
```

For every indirect effect the table reports the first path coefficient (`a`),
the remainder of the product (`b`), the mediated effect size `es` (`ME`, the
product of all coefficients along the path) and the requested quantity (here the
`power` at `n = 200`).

The power of each indirect effect depends on the test selected. With the joint
significance test (Yzerbyt et al., 2018), two effects that share their weakest
path can require the same sample size even when their other coefficients differ,
because the shared (weaker) coefficient drives the power.

# Correlated variables

Variables can be correlated with the `cor()` directive, written on its own line.
The most common use is two _parallel mediators_ whose residuals are correlated:

```{r}
res <- pamlj::pamlmed(
  mode  = "medmodels",
  code  = "m1 ~ .4*x
           m2 ~ .5*x
           y  ~ .1*x + .2*m1 + .3*m2
           cor(m1,m2) = .3",
  aim   = "power",
  n     = 200,
  test  = "joint",
  inspect_cors = TRUE
)
res$powertab$asDF[, c("effect", "es", "power")]
```

`cor(m1,m2) = .3` sets the correlation between `m1` and `m2` to `.3`, following
the `lavaan` convention `m1 ~~ .3*m2`. The separator can be either `:` or `=`,
so `cor(m1,m2):.3` is equivalent. Several `cor()` lines can be specified. If the
requested correlations imply an impossible (non positive-definite) covariance
matrix, the function reports it.

Setting `inspect_cors = TRUE` adds a table of the model-implied correlations
among all variables, which is handy to check what actually entered the model:

```{r}
res$implied_cors$asDF
```

# Minimum detectable effect and sensitivity analysis

When `aim = "es"`, `pamlj` solves for the **minimum detectable effect** (MDE):
given a fixed sample size and the desired power, it finds the smallest indirect
effect the design can detect. Because a mediated effect is a _product_ of path
coefficients, the answer depends on _which_ coefficient is allowed to change
while the others stay fixed.

## Choosing the coefficient to vary

By default, `pamlj` works on each indirect effect separately and resizes its
**first edge** (the path leaving the independent variable, the `a` path). This
needs no extra input — just set `aim = "es"`:

```{r}
res <- pamlj::pamlmed(
  mode  = "medmodels",
  code  = "m1 ~ .4*x
           m2 ~ .5*x
           y  ~ .1*x + .2*m1 + .3*m2",
  aim   = "es",
  n     = 250,
  power = .80,
  test  = "joint"
)
res$powertab$asDF[, c("effect", "a", "b", "es", "power")]
```

To target a _specific_ coefficient instead — for instance the `b` path of a
mediator, or a coefficient shared by several indirect effects — give it a
**symbolic label** in the syntax and select it with the `test:` directive. A
label is a letter placed before the value, written `label*value*variable`:

```{r}
res <- pamlj::pamlmed(
  mode  = "medmodels",
  code  = "m1 ~ .4*x
           m2 ~ .5*x
           y  ~ .1*x + a*.2*m1 + .3*m2
           test: a",
  aim   = "es",
  n     = 250,
  power = .80,
  test  = "joint"
)
res$powertab$asDF[, c("effect", "a", "b", "es", "power")]
```

Here the coefficient of `m1 → y` is labelled `a`, and `test: a` (equivalently
`test = a`) tells `pamlj` to find the minimum detectable effect by resizing
_that_ coefficient. If the label is not defined, or the labelled coefficient is
not part of any indirect effect, the analysis falls back to the per-effect
default.

## How the search works

`pamlj` resizes the selected coefficient — keeping every other coefficient at
its specified value — until the power of the indirect effect(s) that travel
_through_ that coefficient reaches the desired level. When the coefficient lies
on more than one indirect path (a shared edge), the search drives the
**weakest** of those effects to the target, so that every affected effect
reaches _at least_ the requested power. Effects that do not pass through the
coefficient are left unchanged.

### The balanced solution

Resizing a single coefficient cannot always reach the requested power. The power
of an indirect effect, seen as a function of one of its coefficients, is
_unimodal_: it rises, reaches a peak, then falls back as the coefficient
approaches 1 (the mediator becomes collinear with its predictor and the other
coefficients' standard errors explode). The other coefficients on the path
therefore **cap** the achievable power: if the requested power sits above that
peak, no value of the single coefficient can deliver it.

When this happens, `pamlj` switches to a **balanced solution**: instead of
moving one coefficient, it grows _all_ the coefficients of the path together to
a common value, which removes the bottleneck and can reach any power below 1.
A message explains what happened, and the solution is recognisable in the table
because the `a` and `b` columns of an affected effect become equal (every edge
on the path is set to the same value):

```{r}
res <- pamlj::pamlmed(
  mode  = "medmodels",
  code  = "m1 ~ .4*x
           m2 ~ .5*x
           y  ~ .1*x + .2*m1 + .3*m2",
  aim   = "es",
  n     = 100,
  power = .90,
  test  = "joint"
)
res$powertab$asDF[, c("effect", "a", "b", "es", "power")]
```

If even the balanced solution cannot reach the target at the given sample size,
the table reports the _largest_ detectable effect together with the maximum
power it yields, and suggests increasing the sample size. In practice the
balanced solution is most useful as a feasibility check: when it appears, the
chosen sample size is too small for the requested power _whatever single
coefficient you adjust_, and the common value shows how large every path would
have to be to make the design work.

# Tests

The free interface supports the same tests as the other mediation models, set
with the `test` argument:

* `"joint"` — joint significance (Yzerbyt et al., 2018), the default and
  recommended method;
* `"sobel"` — the Sobel test (Sobel, 1982);
* `"parametric"` / `"simulation"` — the Monte Carlo confidence-interval methods
  (MacKinnon, Lockwood & Williams, 2004). The number of simulations is set with
  `mcR`, and `set_seed = TRUE` with `seed = <n>` makes the result reproducible.

```{r}
res <- pamlj::pamlmed(
  mode  = "medmodels",
  code  = "m1 ~ .4*x
           m2 ~ .5*x
           y  ~ .1*x + .2*m1 + .3*m2",
  aim   = "power",
  n     = 200,
  test  = "sobel"
)
res$powertab$asDF[, c("effect", "es", "power")]
```

# References

Yzerbyt, V., Muller, D., Batailler, C., & Judd, C. M. (2018). New
recommendations for testing indirect effects in mediational models: The need to
report and test component paths. *Journal of Personality and Social
Psychology*, 115(6), 929–943.

Sobel, M. E. (1982). Asymptotic confidence intervals for indirect effects in
structural equation models. *Sociological Methodology*, 13, 290–312.

MacKinnon, D. P., Lockwood, C. M., & Williams, J. (2004). Confidence limits for
the indirect effect: Distribution of the product and resampling methods.
*Multivariate Behavioral Research*, 39(1), 99–128.
