---
title: "Power analysis for mixed models, starting from a fitted lmer()/glmer() model"
author: "Marcello Gallucci"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Power analysis for mixed models, starting from a fitted lmer()/glmer() model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
library(pamlj)
library(lme4)
```

`pamlj::pamlmixed()` runs a power analysis for a linear or generalized linear
mixed model. Normally you describe the model by hand, writing a **syntax**
string in which every fixed effect carries an expected coefficient and every
random effect an expected variance (see `?pamlmixed`). This is convenient when
you are planning a study from scratch, but often you already have a fitted
`lme4::lmer()`/`glmer()` model — a pilot study, a previous experiment, or
published data you want to use as the basis for planning a new, adequately
powered study. Retyping that model as a syntax string is tedious and error
prone.

`pamlmixed()` can instead take the fitted model directly, via the `model =`
argument, and extract everything it needs from it.

# A simple example

We use `lme4`'s built-in `sleepstudy` data: reaction time measured on 18
subjects over 10 days of sleep deprivation, with a random intercept and slope
for `Days` by `Subject`.

```{r}
fit <- lmer(Reaction ~ Days + (Days | Subject), data = sleepstudy)
summary(fit)
```

Passing the fitted model to `pamlmixed()` computes the power of the design
actually observed in the fit — the same number of subjects, the same number of
observations per subject, the same fixed-effect coefficients, residual
variance and random-effect (co)variances:

```{r, warning = FALSE}
res <- pamlmixed(model = fit, algo = "mc", mcR = 200, set_seed = TRUE, seed = 1)
res$powertab$asDF
```

That's the whole workflow: fit the model as you normally would, then hand it
to `pamlmixed()`. (`mcR = 200` and `set_seed = TRUE` are used here only to
keep this vignette fast and reproducible; for a real analysis use the default
Monte Carlo settings, or more repetitions, e.g. `mcR = 2000`.)

# What gets extracted automatically

When `model` is supplied, `pamlmixed()` fills in the arguments it would
otherwise need by hand:

* **`syntax`** — the fixed-effect coefficients (`fixef(fit)`) and the
  random-effect variances (from `VarCorr(fit)`), turned into pamlj's model
  syntax.
* **`clusterpars`** — the number of levels and cases per cluster for every
  grouping factor, read off the fit (`ngrps(fit)`, `nobs(fit)`).
* **`categorical`** — any factor predictor, together with the *exact* contrast
  coding it was fitted with (so the extracted coefficients stay correctly
  interpretable — see below).
* **`sigma2`** and **`model_type`** — the residual variance and whether the
  model is linear (`lmer`) or binomial logistic (`glmer`).
* **`aim`** defaults to `"power"` — computing the power of the observed
  design — instead of the usual default `"n"`.

Any of these can still be overridden explicitly; an explicit argument always
wins over what was extracted from the model. `pamlmixed()` also reports, via
`message()`, anything about the extraction worth knowing (e.g. a factor's
contrast coding, or a random-effect covariance that couldn't be preserved) —
suppress these with `verbose = FALSE`, as in the examples above.

# Asking a different question: sample size for a follow-up study

The fitted model tells you the power *of the study you already ran*. Often the
real question is different: how large would a *follow-up* study need to be to
reach a target power? Override `aim` and `power`:

```{r, warning = FALSE}
res <- pamlmixed(model = fit, aim = "n", power = .90,
                  algo = "mc", mcR = 200, set_seed = TRUE, seed = 1)
res$powertab$asDF
```

By default this searches for the number of cases per cluster (`find = "n"`,
i.e. how many measurement occasions per subject); pass `find = "k"` to search
for the number of clusters (subjects) instead.

# Focusing on one specific effect

A model can have several fixed effects, each with its own power; by default
`pamlmixed()` reports (and, for `aim = "n"`, targets) the *worst-case* one. To
focus the search on one specific effect instead, use `focus =`, giving the
term's name exactly as it appears in the fitted model
(`attr(terms(fit), "term.labels")`):

```{r, warning = FALSE}
res <- pamlmixed(model = fit, focus = "Days", aim = "n", power = .90,
                  algo = "mc", mcR = 200, set_seed = TRUE, seed = 1)
res$powertab$asDF
```

# Categorical predictors

Factor predictors are handled automatically. Here we add a 3-level grouping
factor to the fixed effects:

```{r, warning = FALSE}
sleepstudy2 <- sleepstudy
sleepstudy2$grp <- factor(rep(c("A", "B", "C"), length.out = nrow(sleepstudy)))
fit2 <- lmer(Reaction ~ Days + grp + (Days | Subject), data = sleepstudy2)

res <- pamlmixed(model = fit2, algo = "mc", mcR = 200, set_seed = TRUE, seed = 1)
res$powertab$asDF
```

`pamlmixed()` reads the *exact* contrast matrix `grp` was fitted with directly
off the model's data, rather than guessing or re-coding it, so the extracted
coefficients keep meaning the same thing they did in the original fit.

# Two (crossed) grouping factors

Designs with two crossed grouping factors — for instance subjects and items in
a repeated-measures experiment, `(1 | subject) + (1 | item)` — are supported
as well. `pamlmixed()` reproduces the random-effect covariance between
subject/item terms when the fit has one, and, importantly, checks whether a
fixed-effect predictor is *constant within* one of the two grouping factors
(e.g. a property of the item that never changes across subjects). When it is,
it is simulated as a **between-cluster** effect, so the simulated design does
not need to needlessly balloon in size to represent it. A predictor that
genuinely varies within *every* combination of subject and item (a true
trial-level, counterbalanced factor) has no such shortcut, and still needs
enough real replication in the original data to be simulated at all.

```{r, warning = FALSE}
set.seed(1)
nsubj <- 20; nitem <- 20
cond_per_subject <- factor(sample(c("control", "treatment"), nsubj, replace = TRUE))
d <- expand.grid(subject = factor(1:nsubj), item = factor(1:nitem))
d$cond <- cond_per_subject[as.integer(d$subject)]
d$y <- rnorm(nrow(d), mean = ifelse(d$cond == "treatment", .3, 0))

fit3 <- lmer(y ~ cond + (1 | subject) + (1 | item), data = d)
res <- pamlmixed(model = fit3, algo = "mc", mcR = 200, set_seed = TRUE, seed = 1)
res$powertab$asDF
```

# Limitations

* Only Gaussian (`lmer`) and binomial (`glmer`) models are supported.
* Random-effect covariances are preserved for single-value (non-categorical)
  random terms only; a covariance touching a categorical random slope cannot
  be represented and is dropped, with a message.
* A fixed-effect predictor that varies freely within every grouping factor
  (rather than being constant within at least one of them) needs enough real
  replication per combination of grouping factors to be simulated; if the fit
  doesn't have that, model construction fails with an explanatory message.

For the full set of options — correlated random effects, custom contrast
coding, etc. — see `?pamlmixed`.
